Agda Meets Accelerate

### 2013-03-04 paper/main.tex
paper revision, first round

\DefineVerbatimEnvironment{code}{Verbatim}{} % Add fancy options here if you like.
\DefineVerbatimEnvironment{hcode}{Verbatim}{formatcom=\color{blue},frame=leftline} % Add fancy options here if you like.

@@

Dependent types are an emerging approach to certified programming, where invariants are established in the form of types and proven at compile time. Many of Haskell's type-level extensions used in Accelerate approximate aspects of dependently-typed programming. Hence, it is natural to ask whether fully-fledged dependent types, such as those provided by Agda, improve the specification of an embedded language like Accelerate, whether they increase the scope of static guarantees, and whether they may be leveraged to predict performance more accurately.

@@

Accelerate~\cite{ChakravartyKellerLeeMcdonellGrover2011} is a
data-parallel array language embedded into Haskell, which targets
GPUs. It is a \emph{generative library}, as its data-parallel array
operations are not executed directly, but instead construct abstract
syntax trees (AST) representing an entire data-parallel
subcomputation. These \emph{computation representations} are executed
using a \verb+run+ operation that accepts such a representation (of type
\verb+Acc a+), compiles it to GPU kernels, uploads it to a device,
executes it, and retrieves the results.\footnote{To distinguish
  Haskell code from Agda code, we display Haskell code in blue and with
  a vertical bar on the left side.}
%
\begin{hcode}
CUDA.run :: Arrays a => Acc a -> a
\end{hcode}
%
The type class constraint \verb+Arrays a+ restricts the result type to a single array or a tuple of arrays.

@@

The \verb+use+ operation makes a Haskell array available in the
embedded language by wrapping it into the \verb+Acc+ constructor. It
amounts to copying it to GPU memory.\footnote{Accelerate employs a
  caching strategy to avoid the transfer of arrays, which are already
  available in GPU memory.} The operations \texttt{fold} and
\texttt{zipWith} represent collective operations on Accelerate arrays,
effectively producing a representation of an array computation
yielding  a single float value (\texttt{Scalar Float}). The code relies on (type class) overloading: \texttt{0}, \texttt{(+)}, and
\texttt{(*)} are overloaded to construct abstract syntax.

@@

Peebles formalized parts of the API of the Haskell array library Repa
using Agda~\cite{peebles:derpa}. The formalisation relies on the same
shape structure as Accelerate, but array computations are neither
embedded nor can parallel high-performance code be generated.

Swierstra and Altenkirch investigated the use of dependent types for \emph{distributed} array programming~\cite{swierstra-altenkirch:dep-types-for-distr-arrays,swierstra:more-dependent-types}. Their notation for distributed arrays was inspired by X10 and the main focus is on expressing locality aware algorithms.

@@

To demonstrate the Agda embedding in action, we translate the
dot product example from Section~\ref{sec:accelerate} to Agda.
\begin{code}
dotp : forall {E : Elt} {{p : Numeric E}} {n : Nat}
     -> PreVector n E -> PreVector n E -> Scalar E
dotp{E} xs ys = 
  let xs' = use xs
      ys' = use ys
  in  fold _+_ ("0" ::: E) (zipWith _*_ xs' ys')
\end{code}
Unlike the Accelerate code, this function is polymorphic with respect
to the array element type, provided it is numeric.\footnote{In Agda, arguments in double curly braces are \emph{instance arguments}~\cite{DevriesePiessens2011}. We use them much like type class constraints are used in Haskell.} It furthermore
takes the length $n$ as an implicit parameter thus ensuring that the two input
vectors have the same size. The \texttt{PreVector} type of the
arguments corresponds to the plain \texttt{Vector} type in Accelerate,
whereas the result type \texttt{Scalar E} corresponds to \texttt{Acc
  (Scalar E)}---a piece of abstract syntax.

@@

Accelerate's API features expressive type constraints that describe
the shape of the array arguments and results. These constraints ensure
that no shape mismatches occur (e.g., a one-dimensional array cannot be
considered two-dimensional), but they do not ensure at compile
time that the sizes of the dimensions match up. Such a mismatch
results in a run-time error.

@@

Furthermore, functions like \texttt{map} and \texttt{zipWith} obtain
more precise types. The type of \texttt{map} tells us that the input
shape is identical to the output shape:
\begin{verbatim}
map : {A B} {sh} -> (Exp A -> Exp B) -> Array sh A -> Array sh B
\end{verbatim}
Similarly, the type of \texttt{zipWith} restricts its input arrays to
identical shapes:
\begin{verbatim}
zipWith : {A B C} {sh} -> (Exp A -> Exp B -> Exp C)
        -> Array sh A -> Array sh B -> Array sh C
\end{verbatim}
The latter type is more restrictive than the Accelerate implementation
of \texttt{zipWith}. Instead of checking the sizes of the input
arrays, it truncates them to the respective minima. We also developed
an Agda type that directly corresponds to this implementation. It
requires a binary function that computes the minimum of two shapes,
which we leave as an exercise to the reader.

\subsection{Associativity of Operations}
\label{sec:assoc-oper}

@@

In this encoding, filtering is quite simple because the length of
the \texttt{FVector} does not change. Furthermore, filtering could be
extended to multi-dimensional arrays, although it is not clear if an
array that contains non-existing elements or holes is a sensible notion.
\begin{verbatim}
filterF : forall {n : Nat}{E : Elt}
        -> (Exp E -> Exp Bool) -> FVector n E -> FVector n E
filterF {n}{E} pred vec =
  map g vec
  where
  g : Exp (Pair Bool E) -> Exp (Pair Bool E)
  g bx = pair ((fst bx) && p (snd bx)) x
\end{verbatim}
The mapping operation that applies a function to each element of an
array becomes more complicated because it either has to 
materialize a dummy result for each absent element in the argument
vector or apply the function to absent elements, too. This makes \texttt{filter} reminiscent of the \texttt{where} statement of the SIMD language C$^*$~\cite{rose-etal:c-star}.
\begin{verbatim}
mapF : forall {n : Nat}{E F : Elt}
     -> Exp F -> (Exp E -> Exp F) -> FVector n E -> FVector n F
mapF {n}{E}{F} defaultF f vec =
  map g vec
  where
  g : Exp (Pair Bool E) -> Exp (Pair Bool F)
  g bx = if (fst bx) then (pair (fst bx) (f (snd bx)))
                     else (pair (fst bx) defaultF)
\end{verbatim}
On the positive side, some operations can get rid of the absent
elements. In particular, a fold operation which reduces a filtered
vector with a monoid returns a single value. In Accelerate, such a value has type
\texttt{Scalar}, which is a synonym for an array of dimension $0$.
\begin{verbatim}
foldF : forall {n : Nat}{E : Elt}
      -> (Exp E -> Exp E -> Exp E) -> Exp E
      -> FVector n E -> Scalar E
foldF f e vec =
  fold f e (map (\ bx -> if (fst bx) then (snd bx) else e) vec)
\end{verbatim}
Other operations like \texttt{fold1} and the scan operations present
in Accelerate can also be lifted to this representation, but they
retain a notion of absent elements and do not allow to revert to a
non-filtered representation.

@@

We have build an experimental Agda frontend for the Accelerate
language. The goal of this experiment was to explore potential uses of
dependently-typed programming for data-parallel languages.

>>>>

\DefineVerbatimEnvironment{code}{Verbatim}{} % Add fancy options here if you like.
\DefineVerbatimEnvironment{hcode}{Verbatim}{formatcom=\color{blue},frame=single} % Add fancy options here if you like.

@@

Dependent types \cite{MartinLoef1984} are an established approach to certified programming,
where invariants are established in the form of types and proven at
compile time. Many of Haskell's type-level extensions used in
Accelerate approximate aspects of dependently-typed
programming. Hence, it is natural to ask whether fully-fledged
dependent types, such as those provided by Agda, improve the
specification of an embedded language like Accelerate, whether they
increase the scope of static guarantees, and whether they may be
leveraged to predict performance more accurately.

@@

Accelerate~\cite{ChakravartyKellerLeeMcdonellGrover2011} is a
\emph{generative} data-parallel array language embedded into Haskell, which targets
GPUs. Being generative, its data-parallel array
operations are not executed directly. Instead, Accelerate constructs abstract
syntax trees (AST) representing an entire data-parallel
subcomputation. These \emph{computation representations} are executed
using a \verb+run+ operation that accepts such a representation (of type
\verb+Acc a+), compiles it to GPU kernels, uploads it to a device,
executes it, and retrieves the results.\footnote{To distinguish
  Haskell code from Agda code, we display Haskell code in a blue box.}
%
\begin{hcode}
CUDA.run :: Arrays a => Acc a -> a
\end{hcode}
%
The type class constraint \verb+Arrays a+ restricts the result type to a single array or a tuple of arrays.

@@

The \verb+use+ operation makes a Haskell array available in the
embedded language by wrapping it into the \verb+Acc+ constructor. It
amounts to copying it to GPU memory.\footnote{Accelerate employs
  caching to avoid the transfer of arrays that are already
  available in GPU memory.} The operations \texttt{fold} and
\texttt{zipWith} represent collective operations on Accelerate arrays,
effectively producing a representation of an array computation
yielding  a single float value (\texttt{Scalar Float}). The code relies on (type class) overloading: \texttt{0}, \texttt{(+)}, and
\texttt{(*)} are overloaded to construct abstract syntax.

@@

Peebles formalizes parts of the Repa API % of the Haskell array library Repa
using Agda~\cite{peebles:derpa}. The formalisation relies on the same
shape structure as Accelerate, but array computations are neither
embedded nor can parallel high-performance code be generated.

Swierstra and Altenkirch investigated the use of dependent types for
\emph{distributed} array
programming~\cite{swierstra-altenkirch:dep-types-for-distr-arrays,swierstra:more-dependent-types}.
Their notation for distributed arrays is inspired by the X10 language
\cite{Saraswat2009}. They focus on expressing locality awareness.

@@

To demonstrate the Agda embedding in action, we translate the
dot product example from Section~\ref{sec:accelerate} to Agda.
\begin{code}
dotp : forall {E : Elt} {{p : Numeric E}} {n : Nat}
     -> PreVector n E -> PreVector n E -> Scalar E
dotp{E} xs ys = 
  let xs' = use xs
      ys' = use ys
  in  fold _+_ ("0" ::: E) (zipWith _*_ xs' ys')
\end{code}
Unlike the Accelerate code, this function is polymorphic with respect
to the array element type, provided it is numeric.\footnote{In Agda,
  arguments in double curly braces are \emph{instance
    arguments}~\cite{DevriesePiessens2011} that are aggressively
  inferred. We use them like type class constraints in Haskell.}
The length parameter $n$ ensures that the two input
vectors have the same size. The \texttt{PreVector} type of the
arguments corresponds to the plain \texttt{Vector} type in Accelerate,
whereas the result type \texttt{Scalar E} corresponds to \texttt{Acc
  (Scalar E)}---a piece of abstract syntax.

@@

Accelerate's API features expressive type constraints that describe
the shape of the array arguments and results. These constraints ensure
that no shape mismatches occur (e.g., a 1D array cannot be
considered 2D), but they do not ensure at compile
time that the sizes of the dimensions match up. Such a mismatch
results in a run-time error.

@@

Furthermore, functions like \texttt{map} and \texttt{zipWith} obtain
more precise types. The type of \texttt{map} tells us that the input
shape is identical to the output shape:
\begin{verbatim}
map : {A B} {sh} -> (Exp A -> Exp B) -> Array sh A -> Array sh B
\end{verbatim}
Similarly, the type of \texttt{zipWith} restricts its input arrays to
identical shapes:
\begin{verbatim}
zipWith : {A B C} {sh} -> (Exp A -> Exp B -> Exp C)
        -> Array sh A -> Array sh B -> Array sh C
\end{verbatim}
The latter type is more restrictive than the Accelerate implementation
of \texttt{zipWith}. Instead of checking the sizes of the input
arrays, it truncates them to the respective minima. We also developed
an Agda type that directly corresponds to this implementation. It
requires a binary function \texttt{isect} that computes the minimum of
two shapes of the same rank,
which we leave as an exercise to the reader.
\begin{alltt}
zipWith' : \{A B C\} \{shA shB\} \{p : rank shA \(\equiv\) rank shB\}
         -> (A -> B -> C)
         -> Array shA A -> Array shB B -> Array (isect shA shB p) C
\end{alltt}
\subsection{Associativity of Operations}
\label{sec:assoc-oper}

@@

In this encoding, filtering is quite simple because the length of
the \texttt{FVector} does not change. Furthermore, filtering could be
extended to multi-dimensional arrays, although the result might
require careful interpretation.
\begin{verbatim}
filterF : forall {n : Nat}{E : Elt}
        -> (Exp E -> Exp Bool) -> FVector n E -> FVector n E
filterF {n}{E} pred vec = map g vec
  where g : Exp (Pair Bool E) -> Exp (Pair Bool E)
        g bx = pair ((fst bx) && p (snd bx)) x
\end{verbatim}
The mapping operation that applies a function to each element of an
array becomes more complicated because it either has to 
materialize a dummy result for each absent element in the argument
vector or apply the function to absent elements, too. This makes \texttt{filter} reminiscent of the \texttt{where} statement of the SIMD language C$^*$~\cite{rose-etal:c-star}.
\begin{verbatim}
mapF : forall {n : Nat}{E F : Elt}
     -> Exp F -> (Exp E -> Exp F) -> FVector n E -> FVector n F
mapF {n}{E}{F} defaultF f vec = map g vec
  where g : Exp (Pair Bool E) -> Exp (Pair Bool F)
        g bx = if (fst bx) then (pair (fst bx) (f (snd bx)))
                           else (pair (fst bx) defaultF)
\end{verbatim}
Some operations can get rid of absent
elements. A fold operation which reduces a filtered
vector with a monoid returns a single value. In Accelerate, such a value has type
\texttt{Scalar}, which is a synonym for an array of dimension $0$.
\begin{verbatim}
foldF : forall {n : Nat}{E : Elt}
      -> (Exp E -> Exp E -> Exp E) -> Exp E
      -> FVector n E -> Scalar E
foldF f e vec =
  fold f e (map (\ bx -> if (fst bx) then (snd bx) else e) vec)
\end{verbatim}
Operations like \texttt{fold1} and the scan operations
extend to this representation, but they
% retain a notion of absent elements and
cannot revert to a
non-filtered representation.

@@

We have built an experimental Agda frontend for the Accelerate
language. The goal of this experiment was to explore potential uses of
dependently-typed programming for data-parallel languages.
